---
title: "Clustering and annotation"
author: "Bismark Appiah, PhD."
output:
  github_document:
    toc: true
    number_sections: true
    fig_width: 8
    fig_height: 6
---

# Clustering and annotation

**Aim of this section:**  
Identify transcriptionally distinct cell populations and assign biological cell-type identities using marker genes and reference-based annotation.

In this section, we will:

1. Load the pre-processed Seurat object.
2. Select principal components for downstream analysis.
3. Build a cell-cell neighbourhood graph.
4. Identify clusters at different resolutions.
5. Visualise the clusters using UMAP.
6. Compare clustering resolutions and select an appropriate solution.
7. Identify marker genes for each cluster.
8. Assign biological identities manually using marker genes.
9. Compare manual annotation with automated reference-based annotation using Azimuth.

The focus is on understanding **why each step is performed**, rather than simply running the commands.

---

## Load required libraries

The pre-processing and normalisation steps were performed in the previous section. Here we load the packages required for clustering, visualisation, marker identification, and annotation.

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(sctransform)
library(dplyr)
library(ggplot2)
library(patchwork)
library(Azimuth)
```

# Loading the pre-processed data

The previous section performed quality control, filtering, and SCTransform normalisation.

Therefore, we start this section from the pre-processed Seurat object rather than repeating those steps.

```{r load-preprocessed-data}
# Load the pre-processed Seurat object
query <- readRDS("/Users/bismarkappiah/Documents/AA_UCT/clustering_and_annotation/seurat_preprocessed_SCT.rds") #PATH/TO/seurat_preprocessed_SCT.rds
```

> **Important:** Replace `"PATH/TO/seurat_preprocessed_SCT.rds"` with the path to the pre-processed Seurat object on your computer.

We can first inspect the object to confirm that the expected assays, metadata, and variable features are available.

```{r inspect-object}
# Inspect the Seurat object
query

# Check metadata
head(query@meta.data)

# Check available assays
Assays(query)

# Inspect the most variable genes
head(VariableFeatures(query))
```

The object should contain the **SCT assay**, which will be used for the downstream dimensionality-reduction and clustering workflow.

---

# Selecting principal components for downstream analysis

In the previous section, 15 principal components were initially selected based on the elbow plot and inspection of the genes contributing to each PC.

However, the number of PCs used for downstream analysis does not have to be fixed at the first inspection.

It is useful to examine additional PCs and ask whether they contain informative biological signals or whether they increasingly reflect technical variation.

For example, later PCs may contain increased contributions from mitochondrial or other technical genes. The goal is therefore not simply to maximise the number of PCs, but to select a set that captures meaningful biological variation.

## Run PCA

Here we calculate up to 100 PCs so that we can inspect additional components.

```{r run-pca}
query <- RunPCA(
  query,
  assay = "SCT",
  npcs = 100,
  verbose = FALSE
)
```

We can inspect the Seurat object and examine genes contributing to selected PCs.

```{r inspect-pcs}
# Inspect the PCA reduction
print(
  query[["pca"]],
  dims = 15:25,
  nfeatures = 5
)
```

This allows us to inspect PCs beyond the initial 15 and determine whether additional components contain biologically informative genes.

---

## Elbow plot

The elbow plot shows the amount of variation captured by each principal component.

Typically, the first PCs explain more variation, after which the amount of additional variation captured by each PC begins to decrease.

The point where the curve begins to flatten can provide a useful starting point for selecting the number of PCs.

```{r elbow-plot}
ElbowPlot(
  query,
  ndims = 100
)
```

The elbow plot should not be interpreted in isolation. We should also consider the genes contributing to individual PCs and whether they represent meaningful biological variation or technical effects.

For this workshop, we will use the first 20 PCs for downstream analysis.

```{r select-pcs}
# Select PCs for downstream analysis
# Adjust this after examining the PCA results
dims_use <- 1:20
```

> **Workshop note:** The number of PCs is dataset-dependent. In a real analysis, the selected number should be justified using the PCA results, biological signal, and other quality-control information.

---

# Building the cell-cell neighbourhood graph

Before clustering cells, we need to determine which cells are transcriptionally similar to one another.

`FindNeighbors()` identifies cells with similar profiles in PCA space and constructs a **Shared Nearest Neighbor (SNN) graph**.

Conceptually, the graph represents cells as connected points:

```text
Cell A ─── Cell B
  │          │
  │          │
Cell C ─── Cell D
```

Cells with similar transcriptional profiles tend to be connected by more edges.

The resulting graph is then used by `FindClusters()` to identify groups of highly connected cells.

Importantly, clustering is performed using the neighbourhood graph rather than directly from the expression matrix or UMAP coordinates.

```{r neighbourhood-graph}
query <- FindNeighbors(
  query,
  reduction = "pca",
  dims = dims_use
)
```

---

# Finding clusters

`FindClusters()` uses the cell-cell neighbourhood graph generated by `FindNeighbors()` to identify groups of cells with similar neighbourhood structure.

One of the most important parameters is the **resolution**.

Resolution controls the granularity of the clustering:

```text
Lower resolution
      ↓
Fewer, larger clusters

Higher resolution
      ↓
More, smaller clusters
```

For example:

- A lower resolution may identify broad cell populations.
- An intermediate resolution may separate biologically meaningful subpopulations.
- A higher resolution may identify finer cellular states or subtypes.

### Important

Resolution **does not correspond to a fixed number of clusters**.

For example:

```text
resolution = 0.2  → does NOT mean 2 clusters
resolution = 0.5  → does NOT mean 5 clusters
resolution = 1.0  → does NOT mean 10 clusters
```

The number of clusters produced depends on the dataset, neighbourhood graph, number of PCs, and other parameters.

There is therefore no universally "correct" resolution.

The appropriate resolution should ultimately be evaluated using:

- cluster separation,
- marker genes,
- biological knowledge,
- expected cell populations,
- and the purpose of the analysis.

## Initial clustering

We first perform clustering at a low resolution.

```{r initial-clustering}
query <- FindClusters(
  query,
  resolution = 0.05,
  algorithm = 4,
  random.seed = 453
)
```

The random seed is set to make the clustering reproducible.

The particular value `453` is arbitrary. The important point is to use the same seed when reproducing the analysis.

We can examine the number of cells assigned to each cluster.

```{r initial-cluster-sizes}
table(query$seurat_clusters)
```

We can also inspect the available cluster identities.

```{r initial-cluster-identities}
levels(query)
```

---

# Testing different clustering resolutions

Rather than assuming that one resolution is optimal, we can calculate several clustering solutions.

Here we test:

- 0.1
- 0.2
- 0.3
- 0.5
- 1.0

```{r clustering-resolutions}
query <- FindClusters(
  query,
  resolution = c(
    0.1,
    0.2,
    0.3,
    0.5,
    1
  ),
  algorithm = 4,
  random.seed = 453
)
```

Seurat stores the clustering assignments for the different resolutions in the object's metadata.

We can inspect these metadata columns.

```{r clustering-metadata}
colnames(query@meta.data)
```

You should see columns corresponding to the different clustering resolutions, for example:

```text
SCT_snn_res.0.1
SCT_snn_res.0.2
SCT_snn_res.0.3
SCT_snn_res.0.5
SCT_snn_res.1
```

---

# Computing UMAP

UMAP stands for **Uniform Manifold Approximation and Projection**.

It provides a low-dimensional representation of the data that is easier to visualise.

Here we use the selected PCs as input.

```{r compute-umap}
query <- RunUMAP(
  query,
  dims = dims_use,
  seed.use = 453
)
```

UMAP is useful for visualising the relationships between cells.

However, an important distinction is:

> **UMAP visualises the data; it does not define the clusters.**

The clusters were identified from the neighbourhood graph generated using PCA space.

UMAP is therefore a visual representation of the relationships identified during the analysis.

---

# Comparing clustering resolutions

We can now display the different clustering solutions on the **same UMAP**.

This makes it easier to see how increasing the resolution changes the structure of the clustering.

```{r compare-resolutions, fig.width=12, fig.height=10}
DimPlot(
  query,
  reduction = "umap",
  group.by = c(
    "SCT_snn_res.0.05",
    "SCT_snn_res.0.1",
    "SCT_snn_res.0.2",
    "SCT_snn_res.0.3",
    "SCT_snn_res.0.5",
    "SCT_snn_res.1"
  ),
  label = TRUE,
  repel = TRUE,
  ncol = 3
) &
  NoLegend()
```

When comparing the different resolutions, consider:

- Are there clearly separated populations?
- Are clusters extremely small?
- Does increasing the resolution reveal biologically meaningful subpopulations?
- Are apparently distinct clusters supported by marker genes?
- Does a higher resolution simply divide one biologically homogeneous population?
- Are there clusters that appear to contain mixed populations?

The goal is **not** to choose the resolution that produces the largest number of clusters.

Instead, choose a clustering solution that provides a biologically useful representation of the cellular populations in the dataset.

---

# Selecting a clustering resolution for downstream analysis

For this workshop, we will use resolution **0.3** for downstream analysis.

```{r select-resolution}
# Set the selected clustering solution as the active identities
Idents(query) <- "SCT_snn_res.0.3"
```

We can check the resulting clusters.

```{r selected-cluster-sizes}
table(Idents(query))
```

We can then visualise the selected clustering solution.

```{r selected-clustering, fig.width=8, fig.height=6}
DimPlot(
  query,
  reduction = "umap",
  label = TRUE,
  repel = TRUE
) &
  NoLegend()
```

At this point, we have identified transcriptionally distinct clusters.

The next step is to determine **what biological cell types these clusters represent**.

To do this, we first identify genes that are enriched in each cluster.

---

# Finding cluster marker genes

Marker genes are genes that are more highly expressed in one cluster compared with other cells.

They provide evidence that can help us infer the biological identity of each cluster.

`FindAllMarkers()` performs differential expression analysis for each cluster against the remaining cells.

```{r find-markers, fig.width=16, fig.height=9}
DefaultAssay(query) <- "SCT"

markers <- FindAllMarkers(
  query,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.1
)
```

### Important parameters

`only.pos = TRUE`

Only genes positively enriched in the cluster are returned.

`min.pct = 0.25`

A gene must be detected in at least 25% of cells in either the cluster or the comparison population to be tested.

`logfc.threshold = 0.1`

Genes with an average log2 fold-change below approximately 0.1 are excluded.

The resulting table contains several useful columns, including:

- `avg_log2FC` — strength of enrichment in the cluster
- `pct.1` — proportion of cells expressing the gene in the cluster
- `pct.2` — proportion of cells expressing the gene outside the cluster
- `p_val_adj` — multiple-testing adjusted p-value

---

## Filtering marker genes

For annotation, it can be useful to remove genes that are unlikely to be informative for identifying biological cell types.

Here we remove:

- poorly annotated predicted mouse genes beginning with `Gm`
- `Rik` genes
- unmapped Ensembl identifiers
- ribosomal genes
- mitochondrial genes

```{r filter-markers}
markers_clean <- markers %>%
  dplyr::filter(
    p_val_adj < 0.05,
    !grepl("^Gm", gene),
    !grepl("^Rik|Rik$", gene),
    !grepl("^ENS", gene),
    !grepl("^Rpl|^Rps", gene),
    !grepl("^mt-", gene)
  )
```

We can then select the top five marker genes for each cluster based on average log2 fold-change.

```{r top-markers}
top_markers <- markers_clean %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(
    order_by = avg_log2FC,
    n = 5
  )

top_markers
```

These genes provide an initial overview of the transcriptional characteristics of each cluster.

---

# Visualising cluster marker genes

A heatmap provides a convenient way to visualise the expression of the top marker genes across clusters.

```{r marker-heatmap, fig.width=12, fig.height=10}
DoHeatmap(
  query,
  features = unique(top_markers$gene)
) +
  NoLegend()
```

The heatmap can help us identify groups of genes that are characteristic of particular clusters.

However, annotation should not rely on one gene alone.

Instead, we should consider **combinations of marker genes** and whether they provide a coherent biological interpretation.

---

# Cluster annotation

We can now assign biological identities to the clusters.

There are two complementary approaches:

1. **Manual annotation** based on known marker genes.
2. **Automated reference-based annotation** using a well-annotated reference dataset.

We will first perform manual annotation.

---

## Manual cluster annotation

Cell types can be assigned by examining the expression of known marker genes.

For mouse brain data, different neuronal and glial populations can often be distinguished using combinations of marker genes.

For example:

- **GABAergic / inhibitory neurons:** `Gad2`, `Erbb4`, `Dlx6os1`, `Nxph1`
- **Deep-layer / projection neurons:** `Foxp2`, `Syt6`, `Fezf2`
- **Astrocyte / glial populations:** `Slc39a12`, `Gldc`, `Mertk`
- **Endothelial cells:** `Slco1a4`, `Flt1`, `Adgrl4`
- **Vascular / leptomeningeal cells:** `Slc6a13`, `Aldh1a2`, `Ranbp3l`

Other neuronal populations can be investigated using markers such as:

- `Otof`
- `Glra3`
- `Gsg1l`
- `Ebf2`
- `Ebf3`
- `Tshz2`

A useful way to examine several marker genes simultaneously is with a `DotPlot()`.

```{r manual-marker-annotation, fig.width=12, fig.height=8}
DotPlot(
  query,
  features = c(
    # GABAergic / inhibitory neurons
    "Gad2", "Erbb4", "Dlx6os1", "Nxph1",
    
    # Deep-layer / projection neurons
    "Foxp2", "Syt6", "Fezf2",
    
    # Astrocyte / glial
    "Slc39a12", "Gldc", "Mertk",
    
    # Endothelial cells
    "Slco1a4", "Flt1", "Adgrl4",
    
    # Vascular / leptomeningeal cells
    "Slc6a13", "Aldh1a2", "Ranbp3l",
    
    # Other neuronal populations
    "Otof", "Glra3", "Gsg1l",
    "Ebf2", "Ebf3", "Tshz2"
  ),
  dot.scale = 6
) +
  RotatedAxis()
```

In a DotPlot:

- **Dot size** represents the percentage of cells expressing the gene.
- **Colour** represents the average expression level.

This means that a useful cell-type marker should ideally show strong and relatively specific expression in the relevant cluster.

### Important

Cell types should be identified using **combinations of marker genes**, rather than relying on a single marker.

For example, observing expression of one neuronal marker does not necessarily establish a specific neuronal subtype.

---

# Assigning biological labels

After examining the marker genes, we can assign biological labels to the clusters.

First, check the current cluster identities.

```{r inspect-cluster-identities}
levels(Idents(query))
```

For this workshop dataset, we assign the following labels based on the marker expression observed above.

> **Workshop note:** These labels are dataset-specific and should be adjusted if your clustering or marker expression differs.

```{r assign-biological-labels}
celltype_labels <- c(
  "1"  = "Neurons",
  "2"  = "Neurons",
  "3"  = "GABAergic neurons",
  "4"  = "Astrocyte-like glia",
  "5"  = "Neurons",
  "6"  = "Deep-layer projection neurons",
  "7"  = "Endothelial cells",
  "8"  = "Neurons",
  "9"  = "Deep-layer projection neurons",
  "10" = "Neurons",
  "11" = "VLMCs",
  "12" = "Neurons",
  "13" = "Pericytes",
  "14" = "Projection neurons",
  "15" = "Oligodendrocytes",
  "16" = "Glutamatergic neurons",
  "17" = "Astroglial cells"
)
```

We can rename the cluster identities using these labels.

```{r rename-clusters}
query <- RenameIdents(
  query,
  celltype_labels
)
```

It is useful to save these annotations in the metadata so that they remain available even if we later change the active identities.

```{r save-manual-annotation}
query$manual_annotation <- Idents(query)
```

We can visualise the manually assigned cell types on the UMAP.

```{r manual-annotation-umap, fig.width=10, fig.height=7}
DimPlot(
  query,
  reduction = "umap",
  group.by = "manual_annotation",
  label = TRUE,
  repel = TRUE
) +
  NoLegend()
```

At this stage, the clusters have been assigned biological identities based on marker genes.

However, manual annotation can be subjective, particularly when closely related cell types have overlapping transcriptional profiles.

We can therefore use an automated reference-based method as an additional source of evidence.

---

# Automated cluster annotation

Automated annotation methods compare the query dataset with a reference dataset containing cells with known biological identities.

One such approach is **Azimuth**.

Conceptually:

```text
Query cells
     ↓
Compare with reference
     ↓
Identify similar populations
     ↓
Transfer reference labels
```

Azimuth provides reference-based annotation within the Seurat workflow.

For this dataset, we will use the mouse cortex reference.

```{r automated-annotation}
query <- RunAzimuth(
  query,
  reference = "mousecortexref"
)
```

After running Azimuth, several prediction-related metadata columns are added to the Seurat object.

We can inspect the available metadata.

```{r azimuth-metadata}
colnames(query@meta.data)
```

---

## Visualise broad predicted cell classes

We can visualise the broad predicted classes on the same UMAP.

```{r azimuth-class-umap, fig.width=9, fig.height=7}
DimPlot(
  query,
  reduction = "umap",
  group.by = "predicted.class",
  label = TRUE,
  repel = TRUE
) +
  NoLegend()
```

---

## Visualise predicted subclasses

Azimuth also provides more detailed predicted subclasses.

```{r azimuth-subclass-umap, fig.width=10, fig.height=7}
DimPlot(
  query,
  reduction = "umap",
  group.by = "predicted.subclass",
  label = TRUE,
  repel = TRUE
)
```

The automated predictions provide another way of assessing whether the manually assigned cell identities are consistent with a well-annotated reference dataset.

---

# Automated versus manual annotation

Automated annotation is useful as **supporting evidence**, but it should not replace examination of marker genes and biological interpretation.

A reference-based method may assign the closest available reference label even when:

- the query contains a cell type not represented in the reference,
- two closely related populations are difficult to distinguish,
- the query and reference datasets differ substantially,
- or the biological context differs between datasets.

Therefore, automated labels should always be evaluated against the observed gene expression patterns.

We can compare the manual and Azimuth annotations on the same UMAP.

```{r compare-annotations, fig.width=14, fig.height=7}
DimPlot(
  query,
  reduction = "umap",
  group.by = c(
    "manual_annotation",
    "predicted.subclass"
  ),
  label = TRUE,
  repel = TRUE,
  ncol = 2
)
```

The two plots can be compared to determine where manual and automated annotation agree or disagree.

If there are disagreements, return to the marker genes and examine the relevant clusters in more detail.

---

# Summary

In this section, we performed the following workflow:

```text
Pre-processed Seurat object
          ↓
        PCA
          ↓
 Select informative PCs
          ↓
   FindNeighbors()
          ↓
   FindClusters()
          ↓
Compare resolutions
          ↓
       UMAP
          ↓
 Identify marker genes
          ↓
 Manual annotation
          ↓
 Automated annotation
          ↓
 Compare annotations
```

The key concepts are:

- **PCA** reduces the dimensionality of the expression data and provides the representation used for neighbourhood analysis.
- **FindNeighbors()** constructs a cell-cell similarity graph in PCA space.
- **FindClusters()** identifies groups of highly connected cells in this graph.
- **Resolution** controls clustering granularity and does not correspond to a fixed number of clusters.
- **UMAP** provides a visual representation of the data but does not define the clusters.
- **Marker genes** provide biological evidence for identifying cell populations.
- **Manual annotation** uses known biological markers and expert knowledge.
- **Azimuth** provides complementary reference-based annotation.
- Automated predictions should be interpreted together with marker genes and biological knowledge.

The resulting annotated Seurat object can now be used for downstream analyses such as comparing cell-type abundances, identifying cell-type-specific expression patterns, and investigating biological differences between experimental groups.

# Session information

```{r session-info}
sessionInfo()
```