---
title: "Clustering and annotation"
author: "Bismark Appiah, PhD."
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
    self_contained: true
    toc_float:
      collapsed: false
      smooth_scroll: true
  pdf_document:
    toc: true
---

# Clustering and annotation

**Aim of this section:**  
Identify transcriptionally distinct cell populations and assign biological cell-type identities using marker genes and reference-based annotation

# Install required packages

Run this section once before the workshop.

```{r install-packages, eval=FALSE}
# Install required packages ####################################################
# Run this section once before the workshop
# CRAN packages
cran_packages <- c(
  "Seurat",
  "sctransform",
  "plyr",
  "tidyverse",
  "patchwork",
  "ggrepel",
  "data.table",
  "hdf5r",
  "HGNChelper",
  "scales",
  "openxlsx",
  "remotes"
)

cran_missing <- cran_packages[
  !cran_packages %in% rownames(installed.packages())
]

if (length(cran_missing) > 0) {
  install.packages(cran_missing, dependencies = TRUE)
}


# Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

bioc_packages <- c(
  "SingleCellExperiment",
  "scater"
)

bioc_missing <- bioc_packages[
  !bioc_packages %in% rownames(installed.packages())
]

if (length(bioc_missing) > 0) {
  BiocManager::install(bioc_missing, ask = FALSE, update = FALSE)
}


# GitHub packages
if (!requireNamespace("SeuratWrappers", quietly = TRUE)) {
  remotes::install_github(
    "satijalab/seurat-wrappers",
    upgrade = "never"
  )
}

if (!requireNamespace("Azimuth", quietly = TRUE)) {
  remotes::install_github(
    "satijalab/azimuth",
    ref = "master",
    upgrade = "never"
  )
}
```

# Load libraries

```{r load-libraries, message=FALSE, warning=FALSE}
# Load libraries ###############################################################
if (TRUE){
  library(Seurat)
  library(sctransform)
  library(SeuratWrappers)
  library(SingleCellExperiment)
  library(plyr)
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(readr)  
  library(scater)
  library(openxlsx)
  library(tidyverse)
  library(ggrepel)
  library(data.table)
  library(tibble)
  library(hdf5r)
  library(HGNChelper)
  library(scales)
  library(Azimuth)
}
```

# Pre-processing

Uncomment and run if pre-processing is required.

```{r preprocessing, eval=FALSE}
# Uncomment and run if pre-processing required
# Pre-processing | If needed ##############################################################
if (TRUE){


  # Load 10x data
  counts <- Read10X_h5(
    "filtered_feature_bc_matrix.h5" # Replace with path to your directory
  )

  # Create Seurat object
  seu <- CreateSeuratObject(
    counts = counts,
    project = "singlecell_workshop",
    min.cells = 3,
    min.features = 200
  )


  # Calculate QC metrics
  seu[["percent.mt"]] <- PercentageFeatureSet(
    seu,
    pattern = "^mt-"
  )

  # Look at QC metrics before filtering
  VlnPlot(
    seu,
    features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
    ncol = 3
  )


  # Filter low-quality cells
  seu <- subset(
    seu,
    subset =
      nFeature_RNA > 200 &
      nFeature_RNA < 10000 &
      percent.mt < 5
  )

  # SCTransform
  seu <- SCTransform(
    seu,
    vars.to.regress = "percent.mt",
    verbose = FALSE
  )


  # Compute PCA
  # seu <- RunPCA(
  #   seu,
  #   assay = "SCT",
  #   verbose = FALSE
  # )

  # Save pre-processed object
  # saveRDS(
  #   seu,
  #   file = "PATH_TO_FILE"
  # )

}
```

# Load pre-processed data

```{r load-preprocessed-data}
# Load pre-processed data   ####################################################
seu <- readRDS("~/Documents/AA_UCT/singlecell_workshop/seurat_preprocessed_SCT.rds")

## Selecting PCs for downstream analysis #######################################
#
# In the previous section, 15 PCs were selected based on the
# elbow plot and inspection of the genes contributing to each PC.
# In particular, some later PCs showed increased contribution
# from mitochondrial genes.
#
# However, the number of PCs used for downstream analysis does
# not have to be fixed at the first inspection. We can examine
# additional PCs and ask whether they contain informative
# biological signals.
#
# Here, we inspect PCs beyond the initial 15 to determine whether
# additional components contain useful information.

# Run PCA again and increase the number of principal components
seu <- RunPCA(
  seu,
  assay = "SCT",
  npcs = 100,
  verbose = FALSE
)
# Inspect the object
seu
# Check metadata
head(seu@meta.data)
# Check available assays
Assays(seu)
# Inspect most variable genes
head(VariableFeatures(seu))

# Inspect PCA
print(
  seu[["pca"]],
  dims = 15:25, 
  nfeatures = 5
)


# The elbow plot shows how much variation is captured by each PC.
# PCs before the curve begins to flatten generally contain more
# biological structure and are selected for downstream analysis
# Plot variance contribution

ElbowPlot(
  seu,
  ndims = 100
)

# Choose PCs for downstream analysis
# Adjust after examining the elbow plot
dims_use <- 1:20
```

# Build neighbourhood graph

```{r neighbourhood-graph}
# Build neighbourhood graph ####################################################
# A neighbourhood graph represents cells as connected points, where cells
# with similar gene expression profiles are connected to each other.
#
# FindNeighbors() identifies similar cells in PCA space and constructs
# a Shared Nearest Neighbor (SNN) graph. This graph is then used by
# FindClusters() to identify groups of highly connected cells
seu <- FindNeighbors(
  seu,
  reduction = "pca",
  dims = dims_use
)

```

# Find clusters

```{r find-clusters}
# FindClusters    ##############################################################

# FindClusters() uses the cell-cell neighborhood graph previously constructed by the function "FindNeighbors()" 
# NOTE: It does NOT cluster cells directly from the expression matrix or UMAP coordinates
# The most important parameter to explore is "resolution"
# Resolution controls the granularity of the clustering:
#
#   Lower resolution <- fewer, larger clusters
#   Higher resolution <- more, smaller clusters
#
# For example:
#
#   resolution = 0.2 <- broad cell populations
#   resolution = 0.5 <- intermediate clustering
#   resolution = 1.0 <-  more fine-grained clustering
#
# Note:
# Resolution does NOT correspond to a fixed number of clusters.
# A resolution of 0.5 does not mean 5 clusters
#
# There is also no universally "correct" resolution.
# The appropriate value depends on the dataset and biological
# question. Clusters should ultimately be evaluated using marker
# genes and biological knowledge


# Set a positive random seed for reproducibility
# The specific number is arbitrary; using the same seed helps
# reproduce the same clustering results
seu <- FindClusters(
  seu,
  resolution = 0.05,
  algorithm = 4,
  random.seed = 453
)

# Number of cells per cluster
table(seu$seurat_clusters)

# Inspect current identities
levels(seu)
```

## Test different clustering resolutions

```{r clustering-resolutions}
# Test different clustering resolution #########################################
seu <- FindClusters(
  seu,
  resolution = c(
    0.1,
    0.2,
    0.3,
    0.5,
    1
  ),
  algorithm = 4,
  random.seed = 453
)

# View resulting metadata columns
colnames(seu@meta.data)
```

# Compute UMAP

```{r compute-umap}
# Compute UMAP #################################################################
# UMAP = Uniform Manifold Approximation and Projection
# It reduces high-dimensional data to 2D for visualisation
# Similar cells tend to appear closer together
# UMAP visualises the data; it does not define the clusters
seu <- RunUMAP(
  seu,
  dims = dims_use,
  seed.use = 453
)
```

## Compare clustering resolutions

```{r compare-resolutions}
# Compare clustering resolutions on the same UMAP ##############################
# We will use the same UMAP to compare different clustering
# resolutions.

DimPlot(
  seu,
  reduction = "umap",
  group.by = c(
    "SCT_snn_res.0.05",
    "SCT_snn_res.0.1",
    "SCT_snn_res.0.2",
    "SCT_snn_res.0.3",
    "SCT_snn_res.0.5",
    "SCT_snn_res.1"
  ),
  label = TRUE,
  repel = TRUE,
  ncol = 3
) & NoLegend()
```

## Select clustering resolution for downstream analysis

```{r select-resolution}
# Select clustering resolution for downstream analysis #########################
# After comparing different resolutions, we select resolution 0.3
# for downstream analysis.
#
# Set these cluster assignments as the active cell identities

Idents(seu) <- "SCT_snn_res.0.3"

# Check the selected clusters
table(Idents(seu))

# Visualise the selected clustering
DimPlot(
  seu,
  reduction = "umap",
  label = TRUE,
  repel = TRUE
) & NoLegend()
```

# Find cluster marker genes

```{r, fig.width=16, fig.height=9}
# Find cluster marker genes ####################################################

# FindAllMarkers() identifies genes enriched in each cluster
# compared with the other cells
#
# These marker genes help us infer the biological identity
# of each cluster
# Important columns include:
# avg_log2FC = how strongly the gene is enriched in the cluster
# pct.1 = proportion of cells in the cluster expressing the gene
# pct.2 = proportion of cells outside the cluster expressing the gene
# p_val_adj = multiple-testing adjusted p-value

DefaultAssay(seu) <- "SCT"

markers <- FindAllMarkers(
  seu,
  only.pos = TRUE,
  min.pct = 0.25, # gene is only tested if it is detected in at least 25% of cells in either the cluster or the comparison population
  logfc.threshold = 0.1 # genes with at least about 0.1 log2 fold-change in average expression
)

# Keep informative markers for cell-type annotation
markers_clean <- markers %>%
  dplyr::filter(
    p_val_adj < 0.05,
    !grepl("^Gm", gene),        # predicted mouse genes
    !grepl("^Rik|Rik$", gene),  # starts or ends with Rik
    !grepl("^ENS", gene),       # Ensembl IDs that did not map to conventional gene symbols
    !grepl("^Rpl|^Rps", gene),  # ribosomal genes
    !grepl("^mt-", gene)        # mitochondrial genes
  )
# Look at the top marker genes for each cluster
top_markers <- markers_clean %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 5)

top_markers

# Visualise top markers
DoHeatmap(
  seu,
  features = unique(top_markers$gene)
) + NoLegend()
```

# Cluster annotation - manual and automated

## Manual cluster annotation

```{r manual-marker-annotation}
# Cluster annotation - manual and automated ###########################################################
## Manual cluster annotation ###################################################

# Cell types can be assigned by examining known marker genes
# Below are examples of common markers in mouse brain

# In a DotPlot:
# dot size represents the percentage of cells expressing the gene
# colour represents the average expression level
#
# Cell types are identified from combinations of marker genes
# rather than relying on a single marker.

DotPlot(
  seu,
  features = c(
    # GABAergic / inhibitory neurons
    "Gad2", "Erbb4", "Dlx6os1", "Nxph1",
    
    # Deep-layer / projection neurons
    "Foxp2", "Syt6", "Fezf2",
    
    # Astrocyte / glial
    "Slc39a12", "Gldc", "Mertk",
    
    # Endothelial cells
    "Slco1a4", "Flt1", "Adgrl4",
    
    # Vascular / leptomeningeal cells
    "Slc6a13", "Aldh1a2", "Ranbp3l",
    
    # Other neuronal populations
    "Otof", "Glra3", "Gsg1l",
    "Ebf2", "Ebf3", "Tshz2"
  ),
  dot.scale = 6
) +
  RotatedAxis()
```

## Assign biological labels

```{r assign-biological-labels}
## After examining the marker genes, assign biological labels ##################

# Rename cluster identities based on marker expression.
# These labels should be adjusted after inspecting the DotPlot
# and additional markers where necessary

# Check current cluster identities
levels(Idents(seu))

# Assign labels to clusters
celltype_labels <- c(
  "1"  = "Neurons",
  "2"  = "Neurons",
  "3"  = "GABAergic neurons",
  "4"  = "Astrocyte-like glia",
  "5"  = "Neurons",
  "6"  = "Deep-layer projection neurons",
  "7"  = "Endothelial cells",
  "8"  = "Neurons",
  "9"  = "Deep-layer projection neurons",
  "10" = "Neurons",
  "11" = "VLMCs",
  "12" = "Neurons",
  "13" = "Pericytes",
  "14" = "projection neurons",
  "15" = "Oligodendrocytes",
  "16" = "Glutamatergic neurons",
  "17" = "Astroglial cells"
)

# Rename cluster identities
seu <- RenameIdents(
  seu,
  celltype_labels
)

# Save manual annotation in metadata
seu$manual_annotation <- Idents(seu)

# Visualise manual annotations
DimPlot(
  seu,
  reduction = "umap",
  group.by = "manual_annotation",
  label = TRUE,
  repel = TRUE
) + NoLegend()
```

## Automated cluster annotation

```{r automated-annotation}
## Automated cluster annotation ################################################

# Automated annotation maps cells to a well-annotated reference
# dataset and transfers known cell-type labels to the query cells.
# Each query cell is assigned the label of the most similar
# cell population in the reference dataset
# Azimuth provides reference-based annotation within Seurat


seu <- RunAzimuth(
  seu,
  reference = "mousecortexref"
)

# View available prediction columns
colnames(seu@meta.data)

# Plot broad predicted cell classes
DimPlot(
  seu,
  reduction = "umap",
  group.by = "predicted.class",
  label = TRUE,
  repel = TRUE
) + NoLegend()

# Plot more detailed predicted subclasses
DimPlot(
  seu,
  reduction = "umap",
  group.by = "predicted.subclass",
  label = TRUE,
  repel = TRUE
) 
```

## Automated vs manual annotation

```{r, fig.width=14, fig.height=7}
## Automated vs manual annotation #######################################

# Automated annotation is useful as supporting evidence,
# but predictions should always be checked against marker genes
# and biological knowledge!

# table(
#   Manual = seu$manual_annotation,
#   Azimuth = seu$predicted.subclass
# )

# Compare manual and Azimuth annotations on UMAP
DimPlot(
  seu,
  reduction = "umap",
  group.by = c("manual_annotation", "predicted.subclass"),
  label = TRUE,
  repel = TRUE,
  ncol = 2
)
```

# Session information

```{r session-info}
sessionInfo()
```